A practical look at genetic algorithms

### 2015-12-06 evolutionaryComputation/practicals/report.tex
fixed a bunch of spelling mistakes also added more structere and text

\subsection{Experiment}
After implementing these functions we were tasked to make an experiment of it
and make some graphs of the results and draw your conclusions.

\section{Methodology}
The developing method I chose to use is what I like to call the headless
chicken approach. In here you just start implementing functionalities in
hope the result will be good enough. I think this is preferable to other
methodologies such as PRINCE2 or SCRUM because it has very little
methodology overhead.  With a precise defined assignment like this one I
think reducing overhead is one of the main concerns.
Besides most professional methodologies are more
aimed at making teams works rather than individuals.

@@

It also can use any Java library and integrates seamlessly with Java, (although
calling methods the methods of a Scala project from a Java project can be
a little clumsy sometimes, because Scala supports operator overloading in
a strange way).

Other options I considered were:
\begin{itemize}
	\item Java, which I'm very familiar with but I find quite boring to use.
		Its very verbose.
	\item C\#, the language is good but setting it up is difficult since I
		use Linux, it also has no Gradle support which makes difficult to
		use separate libraries or you have to rely on {.}net libraries only
		which is limited. It also prefers object orientated rather than
		functional programming, Scala prefers functional over object
		orientated. I'm off the opinion that this assignment is quite good
		to try functional programming on. Scala's syntax is also more concise.
		In fact C\# feels pretty dated which it flamboyant use of `;' and the
		strange semantic difference that exists between class brackets and
		method brackets. Scala's type system is also more advanced than
		C#'s, so its in a way more object orientated.
	\item Rust/Go, I'm curious about both these languages, but I'm not at
		all familiar with them, so I probably should do first a little
		hobby project with them before starting doing something serious like
		this, I've already done this with Scala before.
	\item Python, I think the project is in that grey area where weakly
		typed langauges become less powerful in use. Especially considering
		the fact how quickly these algoritms blow up in runtime. Static type
		checking then becomes a gift form heaven.
	\item C++, I don't want to deal with C++.
\end{itemize}

\subsection{Architecture}
%TODO: check spell from here
The code is split up into 4 parts, first there are the case classes. These are
immutable classes that hold the domain information (Java beans).
Then there are the algorithm classes and objects. These are pretty straight
forward logical structures. They implement the atoms of possible GA's, such
as for example uniform crossover. After that comes the experiment and the result
classes, the experiment just uses the GA in a specific way.
Finally there is the presentation object \emph{Main}. the only thing it does
is interpret the experiment into this case into a latex structure. There is no
diagram discussing Main, because there is little architecture involved.

@@

The case classes add meaning to the structures.
A case class is a Scala concept were you just define the attributes. Then
Scala generates the `equals', `hashCode' and `toString' methods for you.
They're also by default immutable, which is something I kept intact. You
can make mutations by using the generated copy function, in which you can
specify what is changed.

A \emph{Member} is something with a gene. The fitness value gets pre calculated by
the factory, so that every members' fitness is only calculated once. This can
be done reliably because the member is immutable.

The members are generated in batches into a population, either by the paired
population or by the Population companion object which provides a
`createOnesAndZeros' function.
A population will also have a function with it to generate new members.
The population class doesn't use it itself, but when its members are paired
up into \emph{Pair}'s. A \emph{PairedPopulation} can use the creation function
of the original population to create new members.

The \emph{PairedPopulation} has also a utility function pairedToPop,
this receives a how, in which it can
insert zipped genes (char by char paired up strings) and from which comes
a char sequence. This how is implemented by one of the crossing methods.
This abstraction was added because the code around the cross function was
always the same.
The resulting string from `how' is then inserted into a member
factory to create a new member (the member factory then allows mutation
or obfuscation of the new genes).

\begin{figure}[ht!]
\centering
\begin{emp}[classdiag](20,20)
Class.evo("Evolution")(
"+valuation: FitnessEvaluator",
"mateSelector:IMateSelector",
"offSpringGenerator: PairedPopulation => Population",
"fittestFilter: (Population, Population) => Population",
"hasGoodEnoughSolution: Population => Boolean")(
"startGenetic(seed:Population): Seq[Population]",
"genetic(prev:Seq[Population]): Seq[Population]",
"step(parents:Population): Population",
);
Instance.fit("Fitness")(
	"uniformlyScaled()",
	"linearlyScaled()",
	"blockValuation(block:Seq[Float])",
	"createProbe(function):FitnessEvaluator"
);
Instance.filt("FittestFilter")(
	"killParents()",
	"truncateElitism()",
	"tournementElitism()"
	);
Instance.select("MateSelection")(
	"tournementSelect()"
);
Instance.factor("MemberFactories")(
	"tightlyLinkded()",
	"randomlyLinked()",
	"withCoinflipMutation(function):function",
);
Instance.off("OffspringGenerator")(
	"uniformCross()",
	"singlePointCross()",
	"twoPointCross()"
);

@@

The evolution class is the common structure of a GA{.} It's very simple and
basically follows these steps:
\begin{enumerate}
	\item Select mates.
	\item Generate offspring.
	\item Select the fittest.
\end{enumerate}

@@

The step method calls almost all given functions in above order except for has good
enough solution. This function is called in genetic, which will continue
calling steps until the population is good enough, or until parents
and children are equal. The population is a case class, so
the generated comparison method is used to check if the children and parents
are equal. \emph{Start genetic} makes a seq of population and calls genetic.

The objects defined above all have functions with the types same named evolution
attributes. So the functions are just basically options you can use to create an
evolution from. There are some exceptions which allow wrapping is the
withCoinFlipMutation function in
memberFactories. This function acts as a decorating function. So you can wrap
it around another function to create a new function.

Fitness contains the evaluation functions. The assignment defined four. However
the block evaluation (deceptive/nondeceptive) can be implemented as the
same function with different values for the indices (which blockvaluation 
does). Interesting is that the blockvaluation in code is shorter than the
mathematical definition.
The create probe function is a utility function that allows the counting of
the fitness calls. The initial design didn't have a \emph{FitnessEvaluator},
but the assignment desired the counting of calls so I just wrapped the
function in a object that increases the count on every call.

The offspringGenerator implements the 'how' part of the pairedPopulations.
The mutation isn't part of the offspringGenerator because this is done by
the member factory.

The fittest filter decides how which of both the parents and children 
continues. The most simple implementation is the Killparents, which
returns the children. Truncate elitism selects the best of both parents
and children. Tournament matches child next to each parent
and decide which one is best. There is no kill children implementation
because nothing would ever change.

MateSelection implements tournament select. This is the only mate selection
algorithm implemented currently.

The member factories are able to create members from strings. These are either
the tightly linked or randomly linked. Then the tightly or randomly linked
functions can also be wrapped in a mutation function,
which will mutate the string before passing it to the factory.
Its an decorator idea, only with functions.

@@

\clearpage

Most of the GA architecture was already in place before the assignment became
available. I did this based on the notes I took in class. When the assignment
became available I was almost completly done with the GA{.} There were some bugs
in my code however and I interperted some things wrong. I also didn't write
anything for an experiment. So I still needed to do that.

@@

Two point cross and single cross both seem to find results in deceptive
random tournament elitism.

\subsection{Hypothesis}
The deceptive trap function will not be solved by any algorithm.
non-deceptive may be more successful. The others will be solved.
I also think mutation will have an important effect and may even
allow deceptive to be solved.

I also think that the provided selection pressure would be about
average. It will probably always find a result.

\subsection{Execution}
Note that I added some extra selection methods and crossing strategies to
the final experiment. I did this because I was curious how they would
compete with the original assignment.

@@

>>>>

\subsection{Experiment}
After implementing these functions we were tasked to make an experiment of it
and make some graphs of the results and make a conclusion based on the results.

\section{Methodology}

@@

It also can use any Java library and integrates seamlessly with Java, (although
calling methods the methods of a Scala project from a Java project can be
a little clumsy sometimes, because Scala supports operator overloading in
a strange way, but its possible).

Other options I considered were:
\begin{itemize}
	\item Java, Java is an industry standard language, but the syntax of
		Java is very verbose. So when I have the choice I try to avoid it.
		I've also used a lot of Java when I was younger so I won't learn
		much by using it.
	\item C\#, the language is good, a lot less verbose then Java
		but setting it up is difficult since I use Linux.
		It also prefers object orientated rather than
		functional programming. Scala's type system is also more advanced than
		C\#'s, so its in a way more object orientated.
	\item Rust/Go, I'm curious about both these languages, but I'm not at
		all familiar with them, so I probably should do first a little
		hobby project with them before starting doing something serious like
		this, this is in contrast to Scala, with which I already played before.
	\item Python, I think the project is in that grey area where weakly
		typed languages become less powerful in use. Especially considering
		the fact how quickly these algorithms blow up in runtime. Static type
		checking then helps solving a big part of the bugs before execution
		and since execution will be slow, this saves a lot of time.
	\item C++, C++ is extremely fast in execution times. However implementation
		time is quite long compared to almost everything else.
		Errors are for example reported poorly and
		the syntax is inconsistent. Since execution speed isn't a top priority
		I decided to not use C++.
\end{itemize}

\subsection{Architecture}
The code is split up into 4 parts, first there are the case classes. These are
immutable classes that hold the domain information (Java beans).
Then there are the algorithm classes and objects. These are pretty straight
forward logical structures. They implement the atoms of possible GA's, such
as for example uniform crossover. After that comes the experiment and the result
classes, the experiment just uses the GA in a specific way.
Finally there is the presentation object \emph{Main}. The only thing it does
is interpret the experiment into a latex structure. There is no
diagram discussing Main, because there is little architecture involved.

\subsubsection{Case classes}
The case classes add meaning to the structures.
A case class is a Scala concept were you just define the attributes. Then
Scala generates the `equals', `hashCode' and `toString' methods for you.
They're also by default immutable, which is something I like using. You
can make mutations by using the generated copy function, in which you can
specify what is changed to create a new copy with the change.

@@

\paragraph{Member} A \emph{Member} is something with a gene. The fitness value gets pre calculated by
the factory, so that every members' fitness is only calculated once. This can
be done reliably because the member is immutable. There is no way to modify
the genes so the fintness value does not have to change.

\paragraph{Population} The members are generated in batches and stored into a
\emph{Population}, either by the paired
population or by the Population companion object which provides a
`createOnesAndZeros' function.
A population will also have a function with it to generate new members as
data in this case, it drags the specified function along, its like a
function pointer.
The population class doesn't use it itself, but when its members are paired
up into \emph{Pair}s. A \emph{PairedPopulation} can use the creation function
of the original population to create new members.

\paragraph{PairedPopulation}
The \emph{PairedPopulation} has also a utility function pairedToPop,
this receives a how, in which it can
insert zipped genes (char by char paired up strings) and from which comes
a char sequence. This how is implemented by one of the crossing methods.
This abstraction was added because the code around the cross function was
always the same.
The resulting string from `how' is then inserted into a member
factory to create a new member (the member factory then allows mutation
or obfuscation of the new genes).

I've thought about merging PairedPopulation and population into
one class. However doing that will probably not make things more obvious.
Although it does makes sense in that the population would actually use
the function it drags along.

\newpage
\subsubsection{Algorithm}
The objects defined above all have functions with the types
same named evolution attributes. So the functions are just basically
options you can use to create an
evolution from. There are some exceptions which allow wrapping is the
withCoinFlipMutation function in
memberFactories. This function acts as a decorating function. So you can wrap
it around another function to create a new function.

\begin{figure}[ht!]
\centering
\begin{emp}[classdiag](20,20)
Class.evo("Evolution")(
"+evaluation: FitnessEvaluator",
"mateSelector:IMateSelector",
"offSpringGenerator: PairedPopulation => Population",
"fittestFilter: (Population, Population) => Population",
"hasGoodEnoughSolution: Population => Boolean")(
"startGenetic(seed:Population): Seq[Population]",
"genetic(prev:Seq[Population]): Seq[Population]",
"step(parents:Population): Population",
);
Instance.fit("Evaluation")(
	"uniformlyScaled()",
	"linearlyScaled()",
	"blockValuation(block:Seq[Float])",
	"createProbe(function):FitnessEvaluator"
);
Instance.filt("FittestFilter")(
	"killParents()",
	"truncateElitism()",
	"tournementElitism()"
	);
Instance.select("MateSelection")(
	"tournementSelect()"
);
Instance.factor("MemberFactories")(
	"tightlyLinkded()",
	"randomlyLinked()",
	"withCoinflipMutation(function):function",
);
Instance.off("OffspringGenerator")(
	"uniformCross()",
	"singlePointCross()",
	"twoPointCross()"
);

@@

\paragraph{Evolution}The evolution class is the common structure of a GA(Gentic algorithm){.} 
So creating an instance of Evolution is like creating a new GA{.}

@@

Some of these functions are abstracted away by traits, but still can be seen
as just fucntions.
\subparagraph{Step method}
The step method calls almost all given functions in the follwoing order:

\begin{enumerate}
	\item Select mates.
	\item Generate offspring.
	\item Select the fittest.
\end{enumerate}

\subparagraph{Genetic} hasGoodEnoughSolution is called in genetic,
which will continue calling steps until the population has a good enough
solution, or until parents and children are equal.
The population is a case class, so the generated comparison method is used
to check if the children and parents are equal.
The result of this function is a sequance of population. The first element
is te first generation and the last is the last generation.

\subparagraph{Start genetic} Start genetic makes a seq of the seed population and
calls genetic.

\paragraph{Evaluation} The evaluation object
contains the evaluation functions. The assignment asked to define four.
However the block evaluation (trap functions) can be implemented as the
same function with different values for the indices. This is done by
blockvaluation.

The create probe function is a utility function that allows the counting of
the fitness calls. The initial design didn't have a \emph{FitnessEvaluator},
but the assignment desired the counting of calls so I just wrapped the
function in a object that increases the count on every call. I had to
replace the function prototype in evolution with a trait because coutning
the calls of a function requires some form of state (an increasing variable)
I didn't know how to do that properly with just functions in scala yet,
so I just used a trait instead (trait = interface + state).

\paragraph{Offspring generator} The offspringGenerator implements the `how'
part of the pairedPopulations.  In there are three crossover functions
uniform, single point and two point crossover.

\subparagraph{Uniform crossover}
Uniform crossover checks character by character if they're equal
If not then a random one or zero is used. This is a strange
implementation, but its valid in this case because the only choice is
between 1 and 0, unequal means that there exist in both the father and
mother a one or zero. But not both. So we can just choose one.

\subparagraph{1 point crossover}
Generates a random number between zero and gene length. On the left side 
the father is used and on the right side the mother.

\subparagraph{2 point crossover}
Two point crossover was initially implemented
as first selecting a random value from 0 to gene length. The second would
be from first to gene length. However this is a biased method in that
on of the parents has more chance to be inherited. So
later I would just generate two random values. Between 0 and gene length
Then if the first that was generated is smaller then the second we use
the father as a base string and the mother as the other sequence. Otherwise
the other way around.

However the unbiased method performs structurally worse as can bee seen
in the experiment.

\subparagraph{Mutation}
Mutation isn't part of the offspringGenerator
because this is done by the member factory. It was easier to implement
in the member factory due to the architecture choices.

\paragraph{Fittest filter} The fittest filter decides how which of both
the parents and children continues.
The most simple implementation is the \emph{Killparents}, which
returns the children. \emph{Truncate elitism} selects the best of both parents
and children. \emph{Tournament elitism} matches child next to each parent
and decide which one is best. There is no kill children implementation
because nothing would ever change.

\paragraph{Mate selection} MateSelection implements tournament select.
This is the only mate selection algorithm implemented currently.

\paragraph{Member factories} The member factories are able to create
members from strings. These are either the tightly linked or randomly
linked. Then the tightly or randomly linked
functions can also be wrapped in a mutation function,
which will mutate the string before passing it to the factory.
Its an decorator idea, only with functions.

@@

@@

Two point cross and single cross both seem to find results in deceptive
random tournament elitism.
\subsection{Hypothesis}
The deceptive trap function will not be solved by any algorithm for
the population sizes. I think this because its to desirable to have
at least one 0000 block in the 100 string length.
Non-deceptive will be more successful. The others will be solved.
I also think mutation will have an important effect and may even
allow deceptive to be solved.

I also think that the provided selection pressure would be about
average. It will probably always find a result.
\subsection{Execution}
I added extra selection methods and crossing strategies to
the final experiment. I did this because I was curious how they would
compete with the original assignment.

@@

\paragraph{Abreviations}
\begin{tabular}{ll}
	Abrivation & Full name \\ \toprule
	2X & 2 Point cross \\
	2XB & 2 Point cross biased \\
	2XM & 2 Point cross mutated \\
	2XM & 2 Point cross mutated \\
	UX & uniform cross \\
	UXM & uniform cross mutated \\
	1X & 1 Point cross \\ \bottomrule
\end{tabular}

The biased two point cross will make 2 random values and use the lowest as
the base crossing for the same parent. The ` normal' 2 point cross will make
2 random values. If the first one generated is higher than the other one it
will make the father the mother.

It turns out however that the biased uniform cross is much more effective.
As it solves the deciptive trap function and leaves the `normal' one in
the dust. I actually excpected an improvement when changing the biased
variant into the current vairant. To my shock it did major damage to the
algoritm. To show this I decided to include both.

\subsubsection{Graph observations}
The major drops are caused by an optimization I implemented. After 2 failed
run the stocastic ensurance function will stop. Often this happens when
zooming in, this makes it appear as if there is a local gap, but its just
random chanche and the fact that the bisectional search is zooming in.
